---
title: "COVID-19 Deaths Analysis"
author: "Zheyuan Fan"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---
The purpose of this analysis is to analyze the relationship between COVID-19 death and time, in different coutries or regions.

```{r eval=FALSE}
install.packages("gamm4")
```

```{r, message=FALSE}
library(devtools)
library(mgcv)
library(gamm4)
library(tidyverse)
```

# COVID-19 data

First, plot deaths from COVID-19, so we can visualize the deaths in five different regions.
```{r}
# Load nCOVID-19 data
covid_data <- read_csv("covid_data.csv")

# Plot over time
covid_data %>% 
  filter(country_region %in% c('Hubei','Italy','Iran','South Korea','USA')) %>% 
  na.omit() %>% 
  ggplot(aes(time, dead, color=country_region)) +
  geom_point() +
  theme_minimal()

# Plot from initial death in region
covid_data %>% 
  filter(country_region %in% c('Hubei','Italy','Iran','South Korea','USA')) %>% 
  na.omit() %>% 
  ggplot(aes(timeInt, dead, color=country_region)) +
  geom_point() +
  theme_minimal()

```

Now fit it a GAM resGam with `dead` as the response a smooth on `timeInt` and `country_region` as covariates. 
```{r}
resGam= mgcv::gam(
  dead ~ s(timeInt, pc=0) + country_region, 
  data=covid_data, 
  family=poisson(link='log'))
```

Now we summarize and get the conclusion of the model, then plot it.

```{r}
summary(resGam)
coef(resGam)
plot(resGam)
```
## Conclusion from above
The estimated degrees of freedom for the smooth of `timeInt` is 8.758, we see an edf is much higher than 1(which is 8.758), which means the relationship between deaths and time is not close to linear. We can interpret the coefficients for `country_region`. For example, country_regionAustralia has a coefficient of 0.078, means time has a positive relationship with deaths due to COVID-19 in Australia, one unit of time will cause 0.078 more deaths in Australia.

Next, we fit and plot two more GAMs with the same model but with `k = 50` and `k = 20`.
```{r}
resGam3= mgcv::gam(
  dead ~ s(timeInt, k=50, pc=0) + country_region, data=covid_data, 
  family=poisson(link='log'), method='ML')
plot(resGam3)
gam.check(resGam3)

resGam4 = mgcv::gam(
  dead ~ s(timeInt, k=20, pc=0) + country_region, data=covid_data, 
  family=poisson(link='log'), method='ML')
plot(resGam4)
gam.check(resGam4)
```

```{r, cache=TRUE}
covid_data$timeIntInd = covid_data$timeInt

resGammInd = gamm4::gamm4(
  dead ~ country_region + 
      s(timeInt, k=20, pc=0),
	random = ~ (1|timeIntInd), 
	data=covid_data, family=poisson(link='log'))

plot(resGammInd$gam)
summary(resGammInd$mer)
summary(resGammInd$gam)

```

The plot suggests a trend where we estimate a sharper increase in deaths per day over the first 25 days to a month and then the number decreases from about day 30 onwards. 

```{r}
covid_data_2 <- expand_grid(covid_data$timeInt, covid_data$country_region) %>% 
  as_tibble() %>% 
  rename(timeInt = 1, country_region = 2) %>% 
  distinct() 
  
covid_data_2$predicted <- predict(resGammInd$gam, newdata=covid_data_2, type="response")

#covid_data_3 <- bind_cols(covid_data_2, predicted) %>% 
  #mutate(lower = fit - 2*se.fit, upper = fit + 2*se.fit)

covid_data_2 %>% 
  ggplot(aes(timeInt, predicted, colour=country_region)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~country_region) +
  ggtitle("Predicted deaths over time (time = 0 is first death)")
```
The plot shows the Predicted deaths over time (time = 0 is first death).

## Fit a different model.
Now we fit a different model with time being a random slope.

```{r, cache=TRUE, eval=FALSE}
covid_data$timeSlope = covid_data$timeInt/100

resGammSlope = gamm4::gamm4(
  dead ~ country_region + s(timeInt, k=30, pc=0),
	random = ~(0+timeSlope|country_region) + 
    (1|timeIntInd:country_region), 
  data=covid_data, family=poisson(link='log'))
#save(resGammSlope, file='resGamSlope.RData')
plot(resGammSlope$gam)
summary(resGammSlope$mer)
names(lme4::ranef(resGammSlope$mer))
theRanef = lme4::ranef(resGammSlope$mer, condVar = TRUE)$country_region
(theRanefVec = sort(drop(t(theRanef))))

Dcountry = 'France'
toPredict = expand.grid(
  timeInt = 0:100, 
  country_region = Dcountry)
toPredict$timeSlope = toPredict$timeIntInd = 
  toPredict$timeInt
thePred = predict(resGammSlope$gam, 
                  newdata=toPredict, se.fit=TRUE)

matplot(toPredict$timeInt, 
        exp(do.call(cbind, thePred) %*% Pmisc::ciMat(0.75)), 
        type='l',
        col=c('black','grey','grey'), 
        ylim = c(0, 25))
points(covid_data[covid_data$country_region == Dcountry,c('timeInt','dead')], 
       col='red')
```


# Appendix

1. The COVID-19 data was retrieved from GitHub and the procedure is shown below.
```{r gettingCOVIDdata, eval=FALSE}
install.packages("devtools")
devtools::install_github("GuangchuangYu/nCov2019")

x1 <- nCov2019::load_nCov2019(lang = 'en')

cutoff=3

x2 = by(x1$global, x1$global[,'country', drop=FALSE], 
		function(xx) {
		xx$incidence = diff(c(0, xx$cum_confirm))
		xx$dead = diff(c(0, xx$cum_dead))
		if(any(xx$cum_dead >= cutoff)) {
			cutoffHere = min(xx[xx$cum_dead >= cutoff,'time'], na.rm=TRUE) +1
			xx$timeInt = as.numeric(difftime(xx$time, cutoffHere, units='days'))
			xx = xx[xx$timeInt >= 0, ]
			xx=					xx[,c('time','timeInt','cum_confirm','cum_dead','incidence','dead','country')]
		} else {
			xx = NULL
		}
		xx
	}, simplify=FALSE)

x3 = by(x1$province, x1$province[,'province', drop=FALSE], 
		function(xx) {
		xx$incidence = diff(c(0, xx$cum_confirm))
		xx$dead = diff(c(0, xx$cum_dead))
		colnames(xx) = gsub("province","country", colnames(xx))
		if(any(xx$cum_dead >= cutoff)) {
			cutoffHere = min(xx[xx$cum_dead >= cutoff,'time'], na.rm=TRUE) +1
			xx$timeInt = as.numeric(difftime(xx$time, cutoffHere, units='days'))
			xx = xx[xx$timeInt >= 0, ]
			xx=					xx[,c('time','timeInt','cum_confirm','cum_dead','incidence','dead','country')]
		} else {
			xx = NULL
		}
		xx
	}, simplify=FALSE)
class(x2) = class(x3) = 'list'
x2 = x2[grep('China', names(x2), invert=TRUE)]
x = c(x2, x3)
x$Hubei[x$Hubei$incidence > 4000,c('dead','incidence')] = NA

tidy_data <- compact(x) %>% bind_rows() %>% 
  rename(country_region = country) %>% 
  filter(dead>0)
write_csv(tidy_data, "covid_data.csv")
```
2. Some codes were modified from the assignment of the class ran by Prof. Brown and Prof. Bolton from the  University of Toronto.